---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data


```{r}

ActivityRaw <- read.csv(unzip("activity.zip"))
Activity <- ActivityRaw[!is.na(ActivityRaw$steps),]

```

## What is mean total number of steps taken per day?


```{r}

StepsPerDay <- tapply(Activity$steps,Activity$date,sum)
MeanStepsPerDay <- mean(StepsPerDay, na.rm = TRUE)
MeanStepsPerDay <- format(as.integer(MeanStepsPerDay),big.mark = ",")
MeanStepsPerDay

```

The mean total number of steps taken per day is `r MeanStepsPerDay`.

## What is the average daily activity pattern?

```{r}

AvgStepsPerInterval <- as.ts(tapply(Activity$steps,Activity$interval,mean))
par(xaxt = "n")
ts.plot(AvgStepsPerInterval,type="l",
        main = "Average Steps by Time of Day",
        xlab = "Time of Day",
        ylab = "Average Steps per Interval")
par(xaxt = "s")
axis(1,at=seq(0,288,36), labels=c("12am","3am","6am","9am","12pm","3pm","6pm","9pm","12am"))

```

```{r}

MaxSteps <- max(AvgStepsPerInterval)
MaxInterval <- names(AvgStepsPerInterval[AvgStepsPerInterval==MaxSteps])
MaxSteps <- as.integer(MaxSteps)

PadText <- gsub("^(.*)","000\\1",MaxInterval)
MaxTime <- gsub(".*(..)(..)$","\\1:\\2",PadText)

```

The maximum average number of steps in a 5 minute interval is <strong>`r MaxSteps`</strong>

This maximum occured at <strong>`r MaxTime`</strong>

## Imputing missing values

```{r}

NAIntervals <- is.na(ActivityRaw$steps)
NumberOfNAs <- format(sum(NAIntervals),big.mark = ",")
PercentMissing <- format(mean(NAIntervals)*100,digits = 4)

```

The number of rows in the data set with missing values is <strong>`r NumberOfNAs`</strong>
or <strong>`r PercentMissing`%</strong>.

These missing values will be replaced by the average of all other values in the same
time interval.

```{r}

ActivityFix <- ActivityRaw

ActivityFix$steps[NAIntervals] <- AvgStepsPerInterval[ActivityFix$interval[NAIntervals]]


```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
require(reshape2)
require(ggplot2)

ActivityDayWk <- Activity
ActivityDayWk$date <- as.Date(ActivityDayWk$date)

WeekendDays <- weekdays(ActivityDayWk$date) %in% c("Saturday","Sunday")
ActivityDayWk$WeekPart[WeekendDays] <- "weekend"
ActivityDayWk$WeekPart[!WeekendDays] <- "weekday"
ActivityDayWk$WeekPart <- as.factor(ActivityDayWk$WeekPart)

x <- dcast(ActivityDayWk, interval ~ WeekPart,value.var="steps",mean)
x <- melt(x,id.vars = c("interval"))
ggplot(x,aes(interval,value)) + geom_line() +facet_wrap( ~ variable, ncol = 1)

```

